ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2024-11-13 paper/Manuscript-Data-Package.Rmd
Data paper update

---
title: "ActiveCA: an open data product on active transportation episodes in Canada"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes the data set of the {ActiveCA} R data package. {ActiveCA} contains open data products to obtain impedance functions for active transportation modes in Canada, retrieved from the General Social Survey collections from 1986 to 2015. The package provides data tables on walking and cycling episodes, detailing the trip origins, destinations, and duration of each episode. The origin and destination categories covers a wide variety of locations, such as home, work or school, libraries, museums, restaurants, bars, sports centers, health clinics, place of worship, and others. Addittionally, the package details the respondent's region characteristics, specifying wheter they live in a metropolitan area and their province of residency. {ActiveCA} enables users to calculate the average travel time for each origin-destination combination for each survey year, considering the two active transportation modes: walking and cycling. For each Census Metropolitan and Agglomeration Area, we estimated distance-decay curves (impedance functions) for each destination, transportation mode and year. The package will continue to expand with contributions from the authors and the broader community through requests in the future. {ActiveCA} is freely accessible for exploration and download from the associated Github repository, where the documentation and code involved in creating and manipulating data and all open data products are detailed."

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(ActiveCA)
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(reshape2)
library(rlang)
library(scales)
library(gridExtra)
library(ggpubr)
library(grid)
library(stats)
```

```{r load-data, include = FALSE}
load(paste0(here(), "/data/gss_episodes.Rda"))
load(paste0(here(), "/data/gss_impedances.Rda"))
```

@@

This paper presents the open data product {ActiveCA}. Open data products
(ODPs) are the outcome of a transparent process that transforms raw data
(open or not) into analysis-ready data, in which all stages of
development follow open principles [@arribas-bel2021]. ODPs differ from
general open data due to their high utility, added value and open
availability. The product presented in this document is an `R` data
package that currently consists of processed data tables retrieved from
the Time Use collections of the General Social Surveys (GSS) from 1986
to 2015 [@statisticscanada2024], aimed to obtain impedance functions for
active transportation modes in Canada.

To create this `R` data package, we collected, cleaned and processed the
Time Use collections from the GSS surveys to make them ready for
analysis. An `R` data package contains code, data and documentation in a
standardized format that can be installed by `R` users via a centralized
software repository, such as CRAN (Comprehensive R Archive Network) and
GitHub. Although the GSS surveys are publicly sourced and managed by
Statistic Canada, preparing them for analysis can be time-consuming,
tedious and perhaps not even possible for those who try, due to a lack
of documentation or prior knowledge.

The aim of this paper is to walk readers through the data sets and
invite others to experiment in its uses and applications. {ActiveCA} is
freely available on GitHub for all to install and freely use in the
spirit of open and reproducible research. Although {ActiveCA} was
designed to obtain impedance functions, we admit and hope that its use
can be adopted in various applications that even go beyond the range of
possibilities we have imagined. Not only the data, but also all the code
documenting the processing methodology is available for consultation and
evaluation in its repository. This package contributes to reducing the
barrier to using the information contained in GSS surveys to provide
data-driven decisions in transportation analysis.

@@

The Main File compiles an large array of aggregated data, summarizing
the answers to the questionnaire and derived variables that summarize
the respondents' time use across different activities, locations, and
social interactions. This file documents the time and duration that
respondents allocate to each activity and location. The Main File
provides a overview of daily routines and social dynamics, not focusing
on individual activity episodes. Additionally, this file categorizes
activities into bigger groups and subcategories, facilitating the data's
analytical utility with additional metrics such as total transit time,
duration spent with household members, and counts of activities and
episodes.

@@

The Episode File records detailed data for each activity episode
reported by respondents. The entry includes the start and end times,
duration, location, and accompanying social context, informing when and
where activities occurred and with whom. The file distinguishes itself
by focusing on individual episodes rather than respondents, with the
data structured around the numerous activity instances that compose a
day of the respondent. Although respondent-specific characteristics are
not included within the Episode File, it is possible to link the Main
File and the Episode File by using an identifiable variable present in
both data sets.

# Descriptive statistics

For each year available from the Time Use surveys, we reviewed the
episode files to identify cases with activities listed as walking or
cycling and selected the activities immediately before and after the
mobility episode. The {ActiveCA} package includes all
`r nrow(gss_episodes)` episodes that recorded walking or cycling as a
mode of transportation, with trip durations ranging from 0 to 900
minutes. Among the analyses possible with this package, Table
\ref{tab:table-01} presents descriptive statistics on walking and
cycling trips between 1986 and 2015, including metrics such as the count
of recorded trips (count), and measures of trip duration in minutes:
maximum (max), mean, median, and minimum (min). The 1986 survey did not
include bicycle trips.

```{r table-01, echo=FALSE, cache=FALSE}

# Calculate the count of trips for each mode and year
# Summarize counts by year and mode
episodes_counts <- gss_episodes %>%
  group_by(YEAR, MODE) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = YEAR, values_from = count)

# Calculate min, mean, and max statistics
stats_data <- gss_episodes %>%
  group_by(YEAR, MODE) %>%
  summarise(
    min = min(DURATION),
    mean = mean(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(min, mean, median, max), names_to = "Statistic", values_to = "Value") %>%
  pivot_wider(names_from = YEAR, values_from = Value)

@@

# Generate the table with 'Statistic' as the second column
styled_ch03_table_01 <- final_data %>%
  kable(format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "\\label{tab:table-01}Descriptive statistics for episodes with active transport records",  
        align = c('l', 'l', rep('c', ncol(final_data) - 2))) %>%
  kable_styling(full_width = FALSE, font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 2, "Year" = ncol(final_data) - 2)) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")

@@

```{r table-02, echo=FALSE, cache=FALSE, warning=FALSE, results="asis"}
# Filter for selected years
episodes_stats_before <- gss_episodes %>%
  filter(YEAR < 2005) %>%
  group_by(YEAR, dest_label, MODE) %>%
  summarise(
    count = n(),
    min = min(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  )

# Calculate percentages
total_counts <- episodes_stats_before %>% 
  group_by(YEAR, MODE) %>% 
  summarise(total_count = sum(count), 
            .groups = "drop")

@@

styled_ch03_table_02 <- wide_data %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(wide_data))),
    caption = "\\label{tab:table-02}Comparison of travel statistics by mode and destination: 1986, 1992, 1998",
    col.names = c("Destination", "Mode*","Min*","Med*","Max*", "(%)*", "Min","Med","Max", "(%)", "Min","Med","Max", "(%)")
  ) %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 6) %>%
  # column_spec(1, width = "1cm") %>%  # Set width of 'Dest' column to 1.5cm
  # column_spec(2:14, width = "0.5cm") %>%  # Set width of remaining columns to 0.7cm
  add_header_above(c(" " = 2, "1986" = 4, "1992" = 4, "1998" = 4)) %>%
  row_spec(0, bold = TRUE, align = "c") %>%
  column_spec(6, border_right = TRUE) %>%
  column_spec(10, border_right = TRUE) %>% 
  footnote(general = c("* The symbols used in this table represent the following: `Min` denotes the minimum time to reach the destination; `Max` denotes the maximum time to reach the destination; `(%)` indicates a percentage of the total time to reach the destination; `Med` refers to the median time to reach the destination"), threeparttable = T)

@@

```{r table-03,echo=FALSE, cache=FALSE, warning=FALSE}
# Rename certain 'dest_label' values
gss_episodes_renamed <- gss_episodes %>%
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))

episodes_stats_after <- gss_episodes_renamed %>%
  filter(YEAR > 2000) %>%
  group_by(YEAR, dest_label, MODE) %>%
  summarise(
    count = n(),
    min = min(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  )

# Calculate percentages
total_counts <- episodes_stats_after %>% group_by(YEAR, MODE) %>% summarise(total_count = sum(count), .groups = "drop")
episodes_stats_after <- left_join(episodes_stats_after, total_counts, by = c("YEAR", "MODE"))
episodes_stats_after$Percentage <- round((episodes_stats_after$count / episodes_stats_after$total_count) * 100, 1)

@@

styled_ch03_table_03 <- wide_data %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(wide_data))),
    caption = "\\label{tab:table-03}Comparison of travel statistics by mode and destination: 2005, 2010, 2015",
    col.names = c("Destination", "Mode*","Min*","Med*","Max*", "(%)*", "Min","Med","Max", "(%)", "Min","Med","Max", "(%)")
  ) %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                font_size = 6) %>%
  column_spec(1, width = "2cm") %>%  # Set width of 'Dest' column to 1.5cm
  column_spec(2:14, width = "0.4cm") %>%  # Set width of remaining columns to 0.7cm
  add_header_above(c(" " = 2, "2005" = 4, "2010" = 4, "2015" = 4)) %>%
  row_spec(0, bold = TRUE, align = "c") %>%
  column_spec(6, border_right = TRUE) %>%
  column_spec(10, border_right = TRUE) %>% 
  footnote(general = c("* The symbols used in this table represent the following: `Min` denotes the minimum time to reach the destination; `Max` denotes the maximum time to reach the destination; `(%)` indicates a percentage of the total time to reach the destination; `Med` refers to the median time to reach the destination"), threeparttable = T)

@@

{ActiveCA} also enables visual analysis with traditional exploratory
data analysis techniques. Figures \ref{fig:figure-01} and
\ref{fig:figure-02} show walking and cycling trips from 1992 and 2015
through heat maps. These maps use color gradients to represent the
percentage of trips between various origins and destinations, with
darker colors indicating higher percentages and lighter colors
representing less frequent routes. To avoid overwhelming the reader, we
omitted the heat maps for the other years analyzed.

```{=html}
<!---
I`m not surte if this last phrase makes sense in english! "To avoid overwhelming the reader", "Para nao sobrecarregar o leitor" 
-->
```
In 1992, walking trips with `home` as both the origin and destination
made up the majority, accounting for about 30% of all walking trips.
These trips often involved leisure activities, like short walks or dog
walking. Following this, trips from `home` to `work or school` comprised
18% of walking trips. Overall, `home` emerged as a crucial hub, either
as an origin or destination, with only 5% of trips not involving `home.`
By 2015, `home` remained a significant node, but new locations
distributed the proportion of trips to areas not considered in 1992. In
2015, the highest proportion of trips were from `home` to
`work or school` (12%) and vice versa (11%). `home` to `home` accounted
for 8% of trips, and `grocery stores` became a notable destination for
those leaving `home` (6%), surpassing trips to `other's home` (4%).

```{r heatmaps, echo=FALSE, cache=FALSE, include=FALSE}
gss_heatmap <- gss_episodes_renamed %>%
  mutate(WGHT_EPI = ifelse(is.na(WGHT_EPI), 1, WGHT_EPI)) %>% 
  filter(YEAR %in% c(1992,2015))

@@

      # Calculate the total weighted count for normalization
      total_weighted_count <- sum(trip_counts$weighted_count)

@@

      # Define the order of destinations for 2015
      destination_order <- c("Home", "Work or school", "Grocery store", 
                             "Others' home", "Outdoors", "Neighbourhood", 
                             "Cultural venues", "Place of worship", 
                             "Restaurant", "Sport area", "Business")

@@

             )

  }
} 
```

```{r make-figs, include=FALSE}
heatmap_plot_2015_Walking <- heatmap_plot_2015_Walking +
  coord_fixed(ratio = 1) +
  theme(
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),  
    axis.text.x = element_text(size = 14),  
    axis.text.y = element_text(size = 14))

@@

# Save the combined plot to a file
ggsave(file = paste0(here::here(),"/Data-Paper/Manuscript-figures/walking_hm_fig.jpg"), 
       plot = walking_hm_fig,
       width = 17, 
       height = 8, 
       units = "in", 
       dpi = 300)

@@

# Save the combined plot to a file
ggsave(file = paste0(here::here(),"/Data-Paper/Manuscript-figures/cycling_hm_fig.jpg"), 
       plot = cycling_hm_fig,
       width = 17, 
       height = 8, 
       units = "in", 
       dpi = 300)
```

@@

knitr::include_graphics(paste0(here::here(),"/Data-Paper/Manuscript-figures/walking_hm_fig.jpg"))
```

@@

knitr::include_graphics(paste0(here::here(),"/Data-Paper/Manuscript-figures/cycling_hm_fig.jpg"))
```

# Impedance functions for Canadians Metropolitan and Census Agglomerations areas

Impedance functions reveals important information about the travel
behavior of the population, by describing the relationship between the
population at an origin and their likelihood or ability to travel to
specific destinations to access opportunities [@soukhov2024].

{ActiveCA} provides calibrated impedance functions for Canadian
Metropolitan and Census Agglomeration areas. For each combination of
year, destination, and transportation mode, we fitted the most suitable
impedance function based on empirical data from the GSS surveys.
{ActiveCA} includes a total of
`r nrow(unique(gss_impedances[,c("YEAR","dest_label","f_name","MODE")]))`
distance-decay functions. These were estimated using the `fitdistrplus`
package [@mullerdutang2015], selecting the distribution with the lowest
Akaike information criterion (AIC) among exponential, gamma, log-normal,
normal, and uniform types.

Table \ref{tab:table-04} shows the best impedance functions for walking
trips to `Work or school` across survey years, where all distributions,
except for a gamma function in 1998, are log-normal. Figure
\ref{fig:figure-03} illustrates these fitted functions (red line)
alongside histograms of empirical travel times (grey bars). As for the
others functions, these examples enable to calculate gravity-based
accessibility measures for active transportation modes across various
destinations and temporal scales in Canadian urban areas.

```{r include = FALSE}
visualize_impedance <- function(df_functions, travel_cost){

@@

    # Nome da nova coluna com o valor de year
    column_name <- paste0("year_", year)

    # lnorm impedance 
    if (name_function == "lnorm") {
      x <- x %>%
        mutate(!!sym(column_name) := dlnorm(t,
                                             meanlog = subset$est_1[1],
                                             sdlog = subset$est_2[1]))

    # Gamma impedance 
    } else if (name_function == "gamma") {

      x <- x %>%
        mutate(!!sym(column_name) := dgamma(t,
                                             shape = subset$est_1[1],
                                             rate = subset$est_2[1]))
    }else if (name_function == "norm") {  
      x <- x %>%
        mutate(!!sym(column_name) := dnorm(t,
                                             mean = subset$est_1[1],
                                             sd = subset$est_2[1]))
    }else if (name_function == "exp") {  
      x <- x %>%
        mutate(!!sym(column_name) := dexp(t,
                                             exp = subset$est_1[1]))
    }else if (name_function == "unif") {  
      x <- x %>%
        mutate(!!sym(column_name) := dunif(t,
                                             min = subset$est_1[1],
                                             max = subset$est_2[1]))
    }
  }

@@

```{r , include=FALSE}

max_time <- 100

for(destino in unique(gss_impedances$orig_label)){

@@

  # Create an empty list to store the plots
  plots <- list()

# Loop through each unique year

  for(year in unique(gss_impedances$YEAR)){

@@

# Use gridExtra to arrange the plots in a 2x3 layout and add overall axis labels

>>>>

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes the open data product {ActiveCA}. {ActiveCA} is an `R` data package that contains analysis-ready data from the General Social Survey cicles regarding active travel in Canada. The package provides data set detailing active trip origins, destinations, and duration, covering a wide variety of locations, such as home, work or school, libraries, museums, restaurants, bars, sports centers, health clinics, place of worship, and others. The package also details the respondent's region characteristics, specifying wheter they live in a metropolitan area and their province of residency. {ActiveCA} also provides pre-estimated impedance functions for active travel for Canadian Metropolitan Area and Census Agglomerations, which can be used in accessibility analysis to calculate the cost of travel between different locations. The package will continue to expand with contributions from the authors and the broader community through requests in the future. {ActiveCA} is freely accessible for exploration and download from the associated Github repository, where the documentation and code involved in creating and manipulating data are detailed."

@@

```{r load-packages, include=FALSE, cache=FALSE}
# Load packages
library(ActiveCA)
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(reshape2)
library(rlang)
library(scales)
library(gridExtra)
library(ggpubr)
library(grid)
library(stats)
```

```{r load-data, include = FALSE}
# Load data
load(paste0(here(), "/data/gss_episodes.Rda"))
load(paste0(here(), "/data/gss_impedances.Rda"))
```

@@

This paper presents the open data product {ActiveCA}. Open data products
(ODPs) are the outcome of a transparent process that transforms raw data
(open or not) into analysis-ready data, in which all stages of
development follow open principles [@arribas-bel2021]. ODPs differ from
general open data due to their high utility, added value and open
availability.

{ActiveCA} is an `R` data package that provides analysis-ready data from the General Social Survey of Canada Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) [@statisticscanada2024] regarding active travel. These data can be used to study walking and cycling in Canada for various purposes over time, and to calculate impedance functions for accessibility analysis. The package provides pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas.

To create this `R` data package, we collected, cleaned and processed the
Time Use collections from the GSS surveys to make them ready for
analysis. An `R` data package contains code, data and documentation in a
standardized format that can be installed by `R` users via a centralized
software repository, such as CRAN (Comprehensive R Archive Network) and
GitHub. Although the GSS surveys are publicly sourced managed by
Statistic Canada [@statisticscanada2024], preparing them for analysis can be time-consuming, tedious and perhaps not even possible for those who try, due to a lack of documentation or prior knowledge.

The aim of this paper is to walk readers through the data sets and
invite others to experiment in its uses and applications. {ActiveCA} is
freely available on GitHub for all to install and freely use in the
spirit of open and reproducible research. In addition of enabling to obtain impedance functions, {ActiveCA} can be adopted in various applications that even go beyond the range of possibilities we have imagined. Not only the data, but also all the code documenting the processing methodology is available for consultation and
evaluation in its repository. This package contributes to reducing the
barrier to using the information contained in GSS surveys to provide
data-driven decisions in transportation analysis.

@@

The main file compiles an large array of aggregated data, summarizing
the answers to the questionnaire and derived variables that summarize
the respondents' time use across different activities, locations, and
social interactions. This file documents the time and duration that
respondents allocate to each activity and location. The Main File
provides a overview of daily routines and social dynamics, not focusing
on individual activity episodes. Additionally, this file categorizes
activities into bigger groups and subcategories, facilitating the data's
analytical utility with additional metrics such as total transit time,
duration spent with household members, and counts of activities and
episodes.

@@

The episode file records detailed data for each activity episode
reported by respondents. The entry includes the start and end times,
duration, location, and accompanying social context, informing when and
where activities occurred and with whom. The file distinguishes itself
by focusing on individual episodes rather than respondents, with the
data structured around the numerous activity instances that compose a
day of the respondent. Although respondent-specific characteristics are
not included within the episode file, it is possible to link the main
file and the episode file by using an identifiable variable present in
both data sets.

# Methodology

For each selected cycle of the GSS surveys, we reviewed the
episode files to identify cases with activities listed as walking or
cycling, selecting the activities immediately before and after the
mobility episode. After that, we labeled the code variables with their appropriate descriptions, identifying each origin and destination, transportation mode, as well the province and urban classification of the respondent's residency.

Additionally, for Census Agglomerations and Census Metropolitan Areas we pre-estimated impedance functions for cycling and walking trips. Impedance functions reveals important information about the travel behavior of the population, by describing the relationship between the population at an origin and their likelihood or ability to travel to specific destinations to access opportunities [@soukhov2024].

For each combination of year, destination, and transportation mode (walking or cycling), we fitted the most suitable impedance function based on empirical data from the GSS surveys. We used the `fitdistrplus` package [@mullerdutang2015] to estimate the functions, selecting the distribution with the lowest Akaike information criterion (AIC) among exponential, gamma, log-normal, normal, and uniform types.

# Results

## Descriptive statistics

Considering GSS Cycles analyzed, we identified `r nrow(gss_episodes)` episodes that recorded active travel episodes, with trip duration ranging from 0 to 900 minutes, to twelve different destinations. {ActiveCA} includes all these episodes ready for analysis. Table \ref{tab:table-01} presents descriptive statistics on walking and cycling trips between 1986 and 2015, including metrics such as the count
of recorded trips (count), and measures of trip duration in minutes:
maximum (max), mean, median, and minimum (min). The 1986 survey did not
include bicycle trips.

```{r table-01, echo=FALSE, cache=FALSE}
# Calculate the count of trips for each mode and year
episodes_counts <- gss_episodes %>%
  group_by(YEAR, MODE) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = YEAR, values_from = count)

# Calculate min, mean, and max statictis (duration)
stats_data <- gss_episodes %>%
  group_by(YEAR, MODE) %>%
  summarise(
    min = min(DURATION),
    mean = mean(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(min, mean, median, max), names_to = "Statistic", values_to = "Value") %>%
  pivot_wider(names_from = YEAR, values_from = Value)

@@

# Creating a latex table
styled_ch03_table_01 <- final_data %>%
  kable(format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "\\label{tab:table-01}Descriptive statistics for episodes with active transport records",  
        align = c('l', 'l', rep('c', ncol(final_data) - 2))) %>%
  kable_styling(full_width = FALSE, font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 2, "Year" = ncol(final_data) - 2)) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")

@@

```{r table-02, echo=FALSE, cache=FALSE, warning=FALSE, results="asis"}
# Filter for selected years and calculate statistics of trip duration
episodes_stats_before <- gss_episodes %>%
  filter(YEAR < 2005) %>%
  group_by(YEAR, dest_label, MODE) %>%
  summarise(
    count = n(),
    min = min(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  )

# Calculate percentages of trips by mode/year 
total_counts <- episodes_stats_before %>% 
  group_by(YEAR, MODE) %>% 
  summarise(total_count = sum(count), 
            .groups = "drop")

@@

# Latex table
styled_ch03_table_02 <- wide_data %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(wide_data))),
    caption = "\\label{tab:table-02}Comparison of travel statistics by mode and destination: 1986, 1992, 1998",
    col.names = c("Destination", "Mode*","Min*","Med*","Max*", "(%)*", "Min","Med","Max", "(%)", "Min","Med","Max", "(%)")
  ) %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 6) %>%
  # column_spec(1, width = "1cm") %>%  # Set width of 'Dest' column to 1.5cm
  # column_spec(2:14, width = "0.5cm") %>%  # Set width of remaining columns to 0.7cm
  add_header_above(c(" " = 2, "1986" = 4, "1992" = 4, "1998" = 4)) %>%
  row_spec(0, bold = TRUE, align = "c") %>%
  column_spec(6, border_right = TRUE) %>%
  column_spec(10, border_right = TRUE) %>% 
  footnote(general = c("* The symbols used in this table represent the following: `Min` denotes the minimum time to reach the destination; `Max` denotes the maximum time to reach the destination; `(%)` indicates a percentage of the total time to reach the destination; `Med` refers to the median time to reach the destination"), threeparttable = T)

@@

```{r table-03,echo=FALSE, cache=FALSE, warning=FALSE}
# Renaming 'dest_label' and 'orig_label' values to reduce table space
gss_episodes_renamed <- gss_episodes %>%
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))

# Selecting surveys after 2000 and calculating trip duration statistics 
episodes_stats_after <- gss_episodes_renamed %>%
  filter(YEAR > 2000) %>%
  group_by(YEAR, dest_label, MODE) %>%
  summarise(
    count = n(),
    min = min(DURATION),
    median = median(DURATION),
    max = max(DURATION),
    .groups = "drop"
  )

# Calculate percentages of trips by mode/year
total_counts <- episodes_stats_after %>% group_by(YEAR, MODE) %>% summarise(total_count = sum(count), .groups = "drop")
episodes_stats_after <- left_join(episodes_stats_after, total_counts, by = c("YEAR", "MODE"))
episodes_stats_after$Percentage <- round((episodes_stats_after$count / episodes_stats_after$total_count) * 100, 1)

@@

# Latex table
styled_ch03_table_03 <- wide_data %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(wide_data))),
    caption = "\\label{tab:table-03}Comparison of travel statistics by mode and destination: 2005, 2010, 2015",
    col.names = c("Destination", "Mode*","Min*","Med*","Max*", "(%)*", "Min","Med","Max", "(%)", "Min","Med","Max", "(%)")
  ) %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                font_size = 6) %>%
  column_spec(1, width = "2cm") %>%  # Set width of 'Dest' column to 1.5cm
  column_spec(2:14, width = "0.4cm") %>%  # Set width of remaining columns to 0.7cm
  add_header_above(c(" " = 2, "2005" = 4, "2010" = 4, "2015" = 4)) %>%
  row_spec(0, bold = TRUE, align = "c") %>%
  column_spec(6, border_right = TRUE) %>%
  column_spec(10, border_right = TRUE) %>% 
  footnote(general = c("* The symbols used in this table represent the following: `Min` denotes the minimum time to reach the destination; `Max` denotes the maximum time to reach the destination; `(%)` indicates a percentage of the total time to reach the destination; `Med` refers to the median time to reach the destination"), threeparttable = T)

@@

{ActiveCA} also enables visual analysis of active travel in Canada using traditional exploratory data analysis techniques. Figures \ref{fig:figure-01} and
\ref{fig:figure-02} show walking and cycling trips from 1992 and 2015
through heat maps. These maps use color gradients to represent the
percentage of trips between various origins and destinations, with
darker colors indicating higher percentages and lighter colors
representing less frequent routes. To avoid overwhelming the reader, we
omitted the heat maps for the other years analyzed.

```{=html}
<!---
I`m not surte if this last phrase makes sense in english! "To avoid overwhelming the reader", "Para nao sobrecarregar o leitor" 
-->
```

In 1992, walking trips with `home` as both the origin and destination
made up the majority, accounting for about 30% of all walking trips.
These trips often involved leisure activities, like short walks or dog
walking. Following this, trips from `home` to `work or school` comprised
18% of walking trips. Overall, `home` emerged as a crucial hub, either
as an origin or destination, with only 5% of trips not involving `home.`
By 2015, `home` remained a significant node, but new locations
distributed the proportion of trips to areas not considered in 1992. In
2015, the highest proportion of trips were from `home` to
`work or school` (12%) and vice versa (11%). `home` to `home` accounted
for 8% of trips, and `grocery stores` became a notable destination for
those leaving `home` (6%), surpassing trips to `other's home` (4%).

```{r heatmaps, echo=FALSE, cache=FALSE, include=FALSE}

# Creating heatmaps for the years of 1992 and 2015
gss_heatmap <- gss_episodes_renamed %>%
  mutate(WGHT_EPI = ifelse(is.na(WGHT_EPI), 1, WGHT_EPI)) %>% 
  filter(YEAR %in% c(1992,2015))

@@

      # Calculate the total weighted count
      total_weighted_count <- sum(trip_counts$weighted_count)

@@

      # Define the order of destinations 
      destination_order <- c("Home", "Work or school", "Grocery store", 
                             "Others' home", "Outdoors", "Neighbourhood", 
                             "Cultural venues", "Place of worship", 
                             "Restaurant", "Sport area", "Business")

@@

             ) 
  }
} 
```

```{r make-figs, include=FALSE}
# Fixing the size of the heatmaps 
heatmap_plot_2015_Walking <- heatmap_plot_2015_Walking +
  coord_fixed(ratio = 1) +
  theme(
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),  
    axis.text.x = element_text(size = 14),  
    axis.text.y = element_text(size = 14))

@@

# Save the combined plot to a file
ggsave(file = paste0(here::here(),"/paper/Manuscript-figures/walking_hm_fig.jpg"), 
       plot = walking_hm_fig,
       width = 17, 
       height = 8, 
       units = "in", 
       dpi = 300)

@@

# Save the combined plot to a file
ggsave(file = paste0(here::here(),"/paper/Manuscript-figures/cycling_hm_fig.jpg"), 
       plot = cycling_hm_fig,
       width = 17, 
       height = 8, 
       units = "in", 
       dpi = 300)
```

@@

knitr::include_graphics(paste0(here::here(),"/paper/Manuscript-figures/walking_hm_fig.jpg"))
```

@@

knitr::include_graphics(paste0(here::here(),"/paper/Manuscript-figures/cycling_hm_fig.jpg"))
```

## Pre-estimated impedance functions for active travel in Canada

{ActiveCA} includes a total of
`r nrow(unique(gss_impedances[,c("YEAR","dest_label","f_name","MODE")]))`
pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas. Table \ref{tab:table-04} shows the best impedance functions for walking trips to `Work or school` across survey years, where all distributions, except for a gamma function in 1998, are log-normal. Figure
\ref{fig:figure-03} illustrates these fitted functions (red line)
alongside histograms of empirical travel times (grey bars). As for the
others pre-estimated functions, these examples allow to calculate gravity-based
accessibility measures for active travel across various destinations and temporal scales in Canadian urban areas.

```{r include = FALSE}
# Function to visualize impedance functions
visualize_impedance <- function(df_functions, travel_cost){

@@

    # new colunm with the year information in the name
    column_name <- paste0("year_", year)

    # lnorm distribution 
    if (name_function == "lnorm") {
      x <- x %>%
        mutate(!!sym(column_name) := dlnorm(t,
                                             meanlog = subset$est_1[1],
                                             sdlog = subset$est_2[1]))

    # Gamma distribution 
    } else if (name_function == "gamma") {

      x <- x %>%
        mutate(!!sym(column_name) := dgamma(t,
                                             shape = subset$est_1[1],
                                             rate = subset$est_2[1]))

    # Norm distribution

    }else if (name_function == "norm") {  
      x <- x %>%
        mutate(!!sym(column_name) := dnorm(t,
                                             mean = subset$est_1[1],
                                             sd = subset$est_2[1]))

    # Exponential distribution

    }else if (name_function == "exp") {  
      x <- x %>%
        mutate(!!sym(column_name) := dexp(t,
                                             exp = subset$est_1[1]))

    # Uniform distribution

    }else if (name_function == "unif") {  
      x <- x %>%
        mutate(!!sym(column_name) := dunif(t,
                                             min = subset$est_1[1],
                                             max = subset$est_2[1]))
    }
  }

@@

```{r , include=FALSE}
# Setting the maximum time to be considered to plot the impedance graphs
max_time <- 100

# Loop to generate all impedance plots at once 
for(destino in unique(gss_impedances$orig_label)){

@@

  plots <- list()

# Loop through each unique year
  for(year in unique(gss_impedances$YEAR)){

@@

# Use gridExtra to arrange the plots in accordance with the number of survey years (some destination exits only in specific surveys)
